Live Functional Programming with Typed Holes

### 2018-07-17 calculus.tex
added ifarxiv flag and decreased image file sizes to comply with arxiv reqs (but arxiv still doesnt like it...)

For the familiar forms of the lambda calculus, the rules again follow prior work.
%
For simplicity, the core calculus includes only a single base type~$b$ with a single constant~$c$, governed by Rule \rulename{SConst} (i.e. $b$ is the unit type).
%
%\matt{Extraneous and uninteresting:}
By contrast, \citet{popl-paper} instead defined a number type with a single operation. That paper also defined sum types as an extension to the core calculus. We follow suit on both counts in the supplemental appendix.%Appendix~\ref{sec:extensions}.
%

@@

Substitution, written $[d/x]d'$, operates in the standard capture-avoiding manner~\cite{pfpl} (see the supplemental appendix for the full definition).
% Appendix~\ref{sec:additional-defns}).
%
The only cases of special interest arise when substitution reaches a hole closure:
\[
\begin{array}{rcl}
  [d/x]\dehole{u}{\sigma}{} & = & \dehole{u}{[d/x]\sigma}{} \\%
  \substitute{d}{x}{\dhole{d'}{u}{\sigma}{}} & = & \dhole{[d/x]d'}{u}{[d/x]\sigma}{}
\end{array}
\]
In both cases, we write~$[d/x]\sigma$ to perform substitution on each expression in the hole environment~$\sigma$, i.e. the environment ``records'' the substitution.
%
For example, $\stepsToD{}
    {\hap{(\halam{x}{b}{\halam{y}{b}{\dehole{u}{[x/x, y/y]}{}}})}{c}}
    {\halam{y}{b}{\dehole{u}{[c/x, y/y]}{}}}$.
%
Beta reduction can duplicate hole closures.
%
Consequently, the environments of different closures with the same hole name may differ,
e.g., when a reduction applies a function with a hole closure body multiple times as in Fig.~\ref{fig:grades-example}.
Hole closures may also appear within the environments of other hole
closures, giving rise to the closure paths described in
Sec.~\ref{sec:paths}.

@@

\begin{thm}[Preservation]
  If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ and
  $\stepsToD{\Delta}{\dexp}{\dexp'}$ then
  $\hasType{\Delta}{\emptyset}{\dexp'}{\htau}$.
\end{thm}
\noindent
%
The proof relies on an analogous preservation lemma for instruction
transitions and a standard substitution lemma stated in
the supplemental appendix.
% Appendix~\ref{sec:additional-defns}.
%
Hole closures can disappear during evaluation, so we rely on simple weakening
of $\Delta$.

The progress theorem establishes that the dynamic semantics accounts
for every well-typed term, i.e. that we have not forgotten some
necessary rules or premises.
%
\begin{thm}[Progress]
  If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ then either
  (a) $\stepsToD{}{\dexp}{\dexp'}$ or
  (b) $\isBoxedValue{\dexp}$ or
  (c) $\isIndet{\dexp}$.
\end{thm}
\noindent
The key to establishing the progress theorem under a non-empty hole
context is to explicitly account for indeterminate forms,
i.e. those rooted at either a hole closure or a failed cast.
%
The proof relies on canonical forms lemmas stated in the supplemental appendix.
% Appendix~\ref{sec:additional-defns}.

\subsubsection{Complete Programs}
%
Although this paper focuses on running \emph{incomplete} programs, it helps
to know that the necessary machinery does not interfere
with running \emph{complete} programs, i.e. those
with no type or expression holes.
%
% Appendix~\ref{sec:additional-defns}
The supplemental appendix defines the predicates~$\isComplete{e}$, $\isComplete{\htau}$, $\isComplete{\hGamma}$ and~$\isComplete{d}$.
%
Of note, failed casts cannot appear in complete internal expressions.
%
The following theorem establishes that expansion preserves program completeness.

@@

The \Hazel implementation demonstrated in Sec.~\ref{sec:examples} 
includes an unoptimized interpreter, written in OCaml, that implements the semantics as described
in this section, albeit for an \Elm-like language that goes somewhat beyond the simple
calculus here. As with many full-scale systems, there is not currently a formal
specification for the full \Elm{} or \Hazel language, but the supplemental appendix
discusses how the standard approach for deriving a ``gradualized'' version of a 
language construct provides most of the necessary scaffolding \cite{DBLP:conf/popl/CiminiS16}, and provides some examples.

@@

Dynamically, the correctness of fill-and-resume depends on
the following \emph{commutativity} property: if there is some sequence of
steps that go from $d_1$ to $d_2$, then one can fill a hole in these
terms at \emph{either} the beginning or at the end of that step
sequence.
%
We write $\multiStepsTo{\dexp_1}{\dexp_2}$ for the reflexive,
transitive closure of stepping (see the supplemental appendix).
% Appendix~\ref{sec:additional-defns}).
%
\begin{thm}[Commutativity]
  If $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\emptyset}{\dexp_1}{\tau}$
  and $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$ and $\multiStepsTo{\dexp_1}{\dexp_2}$
  then $\multiStepsTo{\instantiate{\dexp'}{u}{\dexp_1}}
                     {\instantiate{\dexp'}{u}{\dexp_2}}$.
\end{thm}
%
Critically, this property relies on the version of
the semantics from \Secref{sec:calculus} where evaluation order is
unspecified (i.e. where we omit the bracketed premises).
%
In general, resuming from $\instantiate{\dexp'}{u}{d_2}$ will not
reduce sub-expressions in the same order as a ``fresh'' left-to-right
reduction sequence starting from $\instantiate{\dexp'}{u}{d_1}$.
%
In other words, this notion of fill-and-resume only works for
languages where evaluation order ``does not matter''.
(There are various standard ways to formalize this intuition.
For the sake of space, we review these in the supplemental appendix.)
% Appendix~\ref{sec:confluence}.)

@@

We describe the proof, which is straightforward but involves a
number of lemmas and definitions, in the supplemental appendix.
%Appendix~\ref{sec:hole-filling}.
%
In particular, care is needed to handle the situation where a
now-filled non-empty hole had taken a step in the original evaluation.

>>>>

For the familiar forms of the lambda calculus, the rules again follow prior work.
%
For simplicity, the core calculus includes only a single base type~$b$ with a single constant~$c$, governed by Rule \rulename{SConst} (i.e. $b$ is the unit type).
%
%\matt{Extraneous and uninteresting:}
By contrast, \citet{popl-paper} instead defined a number type with a single operation. That paper also defined sum types as an extension to the core calculus. We follow suit on both counts in the \appendixName.%Appendix~\ref{sec:extensions}.
%

@@

Substitution, written $[d/x]d'$, operates in the standard capture-avoiding manner~\cite{pfpl} (see the \appendixName for the full definition).
% Appendix~\ref{sec:additional-defns}).
%
The only cases of special interest arise when substitution reaches a hole closure:
\[
\begin{array}{rcl}
  [d/x]\dehole{u}{\sigma}{} & = & \dehole{u}{[d/x]\sigma}{} \\%
  \substitute{d}{x}{\dhole{d'}{u}{\sigma}{}} & = & \dhole{[d/x]d'}{u}{[d/x]\sigma}{}
\end{array}
\]
In both cases, we write~$[d/x]\sigma$ to perform substitution on each expression in the hole environment~$\sigma$, i.e. the environment ``records'' the substitution.
%
For example, $\stepsToD{}
    {\hap{(\halam{x}{b}{\halam{y}{b}{\dehole{u}{[x/x, y/y]}{}}})}{c}}
    {\halam{y}{b}{\dehole{u}{[c/x, y/y]}{}}}$.
%
Beta reduction can duplicate hole closures.
%
Consequently, the environments of different closures with the same hole name may differ,
e.g., when a reduction applies a function with a hole closure body multiple times as in Fig.~\ref{fig:grades-example}.
Hole closures may also appear within the environments of other hole
closures, giving rise to the closure paths described in
Sec.~\ref{sec:paths}.

@@

\begin{thm}[Preservation]
  If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ and
  $\stepsToD{\Delta}{\dexp}{\dexp'}$ then
  $\hasType{\Delta}{\emptyset}{\dexp'}{\htau}$.
\end{thm}
\noindent
%
The proof relies on an analogous preservation lemma for instruction
transitions and a standard substitution lemma stated in
the \appendixName.
% Appendix~\ref{sec:additional-defns}.
%
Hole closures can disappear during evaluation, so we rely on simple weakening
of $\Delta$.

The progress theorem establishes that the dynamic semantics accounts
for every well-typed term, i.e. that we have not forgotten some
necessary rules or premises.
%
\begin{thm}[Progress]
  If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ then either
  (a) $\stepsToD{}{\dexp}{\dexp'}$ or
  (b) $\isBoxedValue{\dexp}$ or
  (c) $\isIndet{\dexp}$.
\end{thm}
\noindent
The key to establishing the progress theorem under a non-empty hole
context is to explicitly account for indeterminate forms,
i.e. those rooted at either a hole closure or a failed cast.
%
The proof relies on canonical forms lemmas stated in the \appendixName.
% Appendix~\ref{sec:additional-defns}.

\subsubsection{Complete Programs}
%
Although this paper focuses on running \emph{incomplete} programs, it helps
to know that the necessary machinery does not interfere
with running \emph{complete} programs, i.e. those
with no type or expression holes.
%
% Appendix~\ref{sec:additional-defns}
The \appendixName defines the predicates~$\isComplete{e}$, $\isComplete{\htau}$, $\isComplete{\hGamma}$ and~$\isComplete{d}$.
%
Of note, failed casts cannot appear in complete internal expressions.
%
The following theorem establishes that expansion preserves program completeness.

@@

The \Hazel implementation demonstrated in Sec.~\ref{sec:examples} 
includes an unoptimized interpreter, written in OCaml, that implements the semantics as described
in this section, albeit for an \Elm-like language that goes somewhat beyond the simple
calculus here. As with many full-scale systems, there is not currently a formal
specification for the full \Elm{} or \Hazel language, but the \appendixName
discusses how the standard approach for deriving a ``gradualized'' version of a 
language construct provides most of the necessary scaffolding \cite{DBLP:conf/popl/CiminiS16}, and provides some examples.

@@

Dynamically, the correctness of fill-and-resume depends on
the following \emph{commutativity} property: if there is some sequence of
steps that go from $d_1$ to $d_2$, then one can fill a hole in these
terms at \emph{either} the beginning or at the end of that step
sequence.
%
We write $\multiStepsTo{\dexp_1}{\dexp_2}$ for the reflexive,
transitive closure of stepping (see the \appendixName).
% Appendix~\ref{sec:additional-defns}).
%
\begin{thm}[Commutativity]
  If $\hasType{\hDelta, \Dbinding{u}{\hGamma'}{\htau'}}{\emptyset}{\dexp_1}{\tau}$
  and $\hasType{\hDelta}{\hGamma'}{\dexp'}{\htau'}$ and $\multiStepsTo{\dexp_1}{\dexp_2}$
  then $\multiStepsTo{\instantiate{\dexp'}{u}{\dexp_1}}
                     {\instantiate{\dexp'}{u}{\dexp_2}}$.
\end{thm}
%
Critically, this property relies on the version of
the semantics from \Secref{sec:calculus} where evaluation order is
unspecified (i.e. where we omit the bracketed premises).
%
In general, resuming from $\instantiate{\dexp'}{u}{d_2}$ will not
reduce sub-expressions in the same order as a ``fresh'' left-to-right
reduction sequence starting from $\instantiate{\dexp'}{u}{d_1}$.
%
In other words, this notion of fill-and-resume only works for
languages where evaluation order ``does not matter''.
(There are various standard ways to formalize this intuition.
For the sake of space, we review these in the \appendixName.)
% Appendix~\ref{sec:confluence}.)

@@

We describe the proof, which is straightforward but involves a
number of lemmas and definitions, in the \appendixName.
%Appendix~\ref{sec:hole-filling}.
%
In particular, care is needed to handle the situation where a
now-filled non-empty hole had taken a step in the original evaluation.

### 2018-07-17 examples-3.tex
added ifarxiv flag and decreased image file sizes to comply with arxiv reqs (but arxiv still doesnt like it...)

At the bottom of the cell in Fig.~\ref{fig:cast-errors}, we have three 
example applications of \li{f}, tupled together for concision. All three are statically
well-typed, again because the hole type is universally consistent. 
The result at the bottom of Fig.~\ref{fig:cast-errors} demonstrates that the first application
of \li{f} is dynamically unproblematic. This allows the programmer confirm that 
the first branch of the conditional operates as intended without the need to 
address the typing problems in the other branch. 
This flexibility is a common motivation for dynamic and gradual languages \cite{Bayne:2011:ASD:1985793.1985864}.

@@

In this small example, it might have only been a small burden for the programmer to provide the intended types in the signature for \li{f}, but there are situations (e.g. during rapid prototyping or a live performance) where the programmer might consider the burden more substantial. This approach ensures that dynamic feedback does not exhibit gaps even when there is a dynamic type error.

>>>>

At the bottom of the cell in Fig.~\ref{fig:cast-errors}, we have three 
example applications of \li{f}, tupled together for concision. All three are statically
well-typed, again because the hole type is universally consistent. 
The result at the bottom of Fig.~\ref{fig:cast-errors} demonstrates that the first application
of \li{f} is dynamically unproblematic. This allows the programmer to confirm that 
the first branch operates as intended without the need to 
address the typing problems in the other branch \cite{Bayne:2011:ASD:1985793.1985864}. 
% This flexibility is a common motivation for dynamic and gradual languages .

@@

\ifarxiv

\else
In this small example, it might have only been a small burden for the programmer to provide the intended types in the signature for \li{f}, but there are situations (e.g. during rapid prototyping or a live performance) where the programmer might consider the burden more substantial. This approach ensures that dynamic feedback does not exhibit gaps even when there is a dynamic type error.
\fi

### 2018-07-17 examples.tex
added ifarxiv flag and decreased image file sizes to comply with arxiv reqs (but arxiv still doesnt like it...)

This section gives an example-driven overview of our  approach as implemented in  \Hazel, a live programming environment being developed by \citet{HazelnutSNAPL}. The \Hazel user interface is based roughly on IPython/Jupyter \cite{PER-GRA:2007}, with results appearing below cells containing code, and the \Hazel language is tracking toward feature parity with \Elm~(\url{elm-lang.org}) \cite{czaplicki2012elm,Elm}, a popular pure functional programming language similar to ``core ML'', with which we assume familiarity. \Hazel is intended initially for use by students and instructors in introductory functional programming courses (where \Elm~ has been successful \cite{DBLP:journals/corr/abs-1805-05125}). 
For the sake of 
exposition, we have post-processed the screenshots in this section after generating them in \Hazel to make use of some ``syntactic and semantic sugar'' from \Elm~that was not available in \Hazel as of this writing, e.g. pattern matching in function arguments, list  notation, and record labels (currently there are only tuples). These conveniences are orthogonal to the contributions of this paper; all of the live programming and user interface features demonstrated in the figures in this section have been implemented.

>>>>

This section gives an example-driven overview of our  approach as implemented in  \Hazel, a live programming environment being developed by \citet{HazelnutSNAPL}. The \Hazel user interface is based roughly on IPython/Jupyter \cite{PER-GRA:2007}, with results appearing below cells containing code, and the \Hazel language is tracking toward feature parity with \Elm~(\url{elm-lang.org}) \cite{czaplicki2012elm,Elm}, a popular pure functional programming language similar to ``core ML'', with which we assume familiarity. \Hazel is intended initially for use by students and instructors in introductory functional programming courses (where \Elm~ has been successful \cite{DBLP:journals/corr/abs-1805-05125}). 
For the sake of 
exposition, we have post-processed the screenshots in this section after generating them in \Hazel to make use of some ``syntactic and semantic sugar'' from \Elm~that was not available in \Hazel as of this writing, e.g. pattern matching in function arguments, list  notation, and record labels (currently there are only tuples). These conveniences are orthogonal to the contributions of this paper; all of the user interface features demonstrated in this section have been implemented.

### 2018-07-17 intro.tex
added ifarxiv flag and decreased image file sizes to comply with arxiv reqs (but arxiv still doesnt like it...)

The anonymous supplement includes, in addition to the Agda mechanization, an appendix that (1) provides some straightforward auxiliary definitions and proofs that were omitted from the paper for concision; and (2) defines some simple extensions to the core calculus (namely, numbers, and sum types), together with a brief exposition on defining further extensions. 
% These are not necessary to communicate the fundamental ideas of the paper, but may be of interest to some readers.

>>>>

\ifarxiv
The appendix 
\else
The anonymous supplement includes, in addition to the Agda mechanization, an appendix that 
\fi
(1) provides some straightforward auxiliary definitions and proofs that were omitted from the paper for concision; and (2) defines some simple extensions to the core calculus (namely, numbers, and sum types), together with a brief exposition on defining further extensions. 
% These are not necessary to communicate the fundamental ideas of the paper, but may be of interest to some readers.
